Finiteness in Cubical Type Theory

### 2020-06-12 sections/introduction.tex
more introduction

\chapter{Introduction}
\section{Foreword}
\todo[inline]{Foreword}
\section{Contributions}
We are interested in constructive notions of finiteness, formalised in Cubical
Type Theory.
In contrast to classical finiteness, in a constructive setting there is a wide
variety of predicates which all have some claim to being the formal
interpretation of ``finiteness''.
Broadly, they vary along two axes: informativeness (i.e. how many properties
of the underlying type we can derive from a proof that it is finite), and
restrictiveness (i.e. how many types conform to its particular variety of
finiteness).
We will explore five of these predicates, proving formally their relations and
implications.

@@

We will then focus on one of the predicates: decidable Kuratowski finiteness.
We will construct a category of decidable Kuratowski-finite types, and show that
it forms a \(\Pi\)-pretopos.

We will then take a brief detour into predicates for countably infinite types,
and compare its landscape to that of the finite predicates.

Finally, we will present a case-study in a practical application of the
predicates for finiteness we have constructed: a library for total proof search.
We will show how our earlier work gives a good theoretical foundation for this
library, and allows for a simple and principled API, and we will demonstrate how
some of the more complex theoretical results allow for a more powerful library
with capabilities not seen in comparable proof search libraries.

\subsection{Finiteness Predicates}
Our first contribution is a cataloguing and characterisation of five finiteness
predicates, shown in figure~\ref{finite-classification}.

While our focus is decidable Kuratowski finiteness, in order to prove the
closure properties that we are interested in we will need to work with the
intermediate predicates as well (though they are also interesting in their own
right).

Since we have a proof-search library in mind, we will have to consider how to
present a good API for constructing proofs of finiteness.
In this vein we will show that some simple base types are finite
(\(\mathbf{Bool}\), \(\top\), \(\bot\), etc), and we will provide functions to
combine finiteness proofs together into more complex types (i.e. \(\times\),
\(\uplus\), \(\Sigma\), \(\Pi\), etc).
As we will see this choice of toolbox for building finiteness proofs actually
has a deep categorical justification, but at this point it makes sense to look
at it as just a suitably general and powerful API.

Our focus in this work is decidable Kuratowski finiteness.
We will characterise it, its strength, and its uses, and in particular we will
show that it forms a \(\Pi\)-pretopos.
Our proofs follow the structure of \cite[Chapters 9, 10]{hottbook} and
\cite{rijkeSetsHomotopyType2015}.
We will explore this concept in more detail later, but crucially we will need to
provide proofs of closure under \(\times\), \(\rightarrow\), etc.

In order to prove these closure properties, we will define and use some
intermediate finiteness predicates, classified in
figure~\ref{finite-classification}.
These predicates are also interesting in their own right, though!

The arrows in figure~\ref{finite-classification} are implication functions
between each predicate.
We will provide a function which inhabits each arrow.
Each unlabelled arrow is an unconditional implication: every manifest Bishop
finite set is cardinal finite (lemma~\ref{manifest-bishop-to-cardinal}), for
instance.
The labelled arrows are strengthening proofs: every manifest enumerable set
\emph{with decidable equality} is split enumerable
(lemma~\ref{manifest-enum-to-split-enum}).
Our most significant result here is the proof that a cardinal finite set with a
decidable total order is manifestly Bishop finite.

\subsection{Countability Predicates}
After the finite predicates, we will briefly look at the infinite countable
types, and classify them in a parallel way to the finite predicates
(section~\ref{infinite}).
\subsection{Search}
All of our work is formalised in Cubical Agda
\cite{vezzosiCubicalAgdaDependently2019}.
We will make mention of the few occasions where the formalisation of some proof
is of interest, but the main place where we will discuss the code is in
section~\ref{search}, where we implement a library for proof search, based on
omniscience and exhaustibility.
This library relies directly on the computational content of univalence, which
allows us to (for instance) have \(\Pi\) types in the domain of the search
space.
The language of a \(\Pi\)-pretopos turns out to be exactly the right language
for constructing ``generator''-like expressions.

The full code of the formalisation is available at
\url{https://github.com/oisdk/finiteness-in-cubical-type-theory}.
\section{Notation and Background}
\subsection{Notation}
We work in Cubical Type Theory \cite{cohenCubicalTypeTheory2016}.
For the various type formers we use the following notation:
\begin{description}
  \item[Type] We use \(\mathbf{Type}\) to denote the universe of (small) types.
    ``Type families'' are functions into \(\mathbf{Type}\).
  \item[0 , 1 , 2] We call the \(\mathbf{0}\), \(\mathbf{1}\), and
    \(\mathbf{2}\) types \(\bot\), \(\top\), and \(\mathbf{Bool}\) respectively.
    The single inhabitant of \(\top\) is tt, and the two inhabitants of
    \(\mathbf{Bool}\) are false and true.
    The ``negation'' of a type, written \(\neg A\), means \(A \rightarrow
    \bot\).
  \item[Dependent Sum and Product] We use \(\Sigma\) and \(\Pi\) for the
    dependent sum and product, respectively.
    The two projections from \(\Sigma\) are called fst and snd.
    In the non-dependent case, \(\Sigma\) can be written as \(\times\), and
    \(\Pi\) as \(\rightarrow\).
  \item[Disjoint Union] We define disjoint union as an inductive type.
    \begin{equation}
      \begin{alignedat}{3}
        A \uplus B \coloneqq & \;
        \text{inl} &: A \rightarrow A \uplus B ; \\
        | & \;  \text{inr} &: B \rightarrow A \uplus B ;
      \end{alignedat}
    \end{equation}
    It is also expressible with only \(\Sigma\):
    \(A \uplus B \simeq \Sigma(x : \mathbf{Bool}) , \text{if } x \text{ then
    } A \text{ else } B \).
  \item[Equalities, equivalences, and paths] We use the symbol \(\coloneqq\)
    for definitions.
    \(\simeq\) will be used for equivalences, and \(\equiv\) for equalities.
    Of course, we know that \((A \simeq B) \simeq (A \equiv B)\) by univalence,
    so the distinction isn't terribly important in usage: we will only use one
    or the other as a suggestion of how we constructed it or how it is to be
    used.
  \item[Lambdas] \todo{Figure out how to describe lambdas}
\end{description}
\subsection{Cubical Type Theory}
Cubical Type Theory \cite{cohenCubicalTypeTheory2016} is a constructive type
theory with an implementation in Cubical Agda
\cite{vezzosiCubicalAgdaDependently2019}.
It allows us to do much of the same theory as in HoTT, but crucially the
univalence ``axiom'' is a \emph{theorem}, giving it computational content.
\begin{definition}[Path Types] \label{path-types}
  The equality type (which we denote with \(\equiv\)) in CuTT is the type of
  Paths\footnotemark.
  The internal structure of paths is largely irrelevant to us here, as we will
  generally treat \(\equiv\) as a black-box equivalence relation with
  substitution and congruence.
\end{definition}

>>>>

\chapter{Introduction}
\section{Foreword}
\todo[inline]{Foreword}
\section{Contributions}
We are interested in constructive notions of finiteness, formalised in Cubical
Type Theory \cite{cohenCubicalTypeTheory2016}.
In this paper we will explore five such notions of finiteness, including their
categorical interpretation, and use them to build a simple proof-search library
facilitated in a fundamental way by univalence.
Along the way we will use the Countdown problem
\cite{huttonCountdownProblem2002} as an example, and provide a program which
produces verified solutions to the puzzle.
We will also briefly examine countability, and demonstrate its parallels and
differences with finiteness.
\subsection{The Varieties of Finiteness}
In Section~\ref{finiteness-predicates} we will explore a number of different
predicates for finiteness.
In contrast to classical finiteness, in a constructive setting there is a wide
variety of predicates which all have some claim to being the formal
interpretation of ``finiteness'' \cite{coquandConstructivelyFinite2010}.
The particular predicates we are interested in are organised in
Figure~\ref{finite-classification}: each arrow in the diagram represents a proof
that one predicate can be derived from another.

@@

These finiteness predicates differ along two main axes: informativeness, and
restrictiveness.
More ``informative'' predicates have proofs which contain extraneous information
other than the finiteness of the underlying type: a proof of split enumerability
(Section~\ref{split-enumerability}), for instance, comes with a strict total
order on the underlying type.
We will prove that a more informative finiteness predicate can be
derived from a less informative one by providing the missing information
(Theorem~\ref{cardinal-to-manifest-bishop}).

The ``restrictiveness'' of a predicate refers to how many types it admits into
its notion of ``finite''.
There are strictly more Kuratowski finite (Section~\ref{kuratowski}) types than
there are Cardinally finite (Section~\ref{cardinal-finiteness}).
We will prove that we can always derive the less restrictive predicate from the
more restrictive one, and that we can go in the other direction by satisfying
the missing requirement (decidable equality in all of these cases).

Proofs coming with extra information is a common theme in constructive
mathematics: often this extra information is in the form of an algorithm which
can do something useful related to the proof itself.
Indeed, our proofs of finiteness here will provide an algorithm to solve the
countdown puzzle.
Occasionally, however, the extra information is undesirable: we may want to
assert the existence of some value \(x : A\) which satisfies a predicate \(P\)
without revealing \emph{which} \(A\) we're referring to.
More concretely, we will need in this paper to prove that two types are in
bijection without specifying a particular bijection.
This facility is provided by Homotopy Type Theory \cite{hottbook} in the form of
propositional truncation, and it is what allows us to prove the bulk of
propositions in this paper.

For each predicate we will also prove its closure properties (i.e. that the
product of two finite sets is finite).
The most significant of these closure proofs is that of closure under \(\Pi\)
(dependent functions) (Theorem~\ref{split-enum-pi}).
\subsection{Toposes and Finite Sets}
In Section~\ref{topos}, we will explore the categorical interpretation of
decidable Kuratowski finite sets.
The motivation here is partially a practical one: by the end of this work we
will have provided a library for proof search over finite types, and the
``language'' of a topos is a reasonable choice for a principled language for
constructing proofs of finiteness in the style of QuickCheck
\cite{claessenQuickCheckLightweightTool2011} generators.

Theoretically speaking, showing that sets in Homotopy Type Theory form a topos
(with some caveats) is an important step in characterising the categorical
implications of Homotopy Type Theory, first proven in
\cite{rijkeSetsHomotopyType2015}.
Our work is a formalisation of this result (and the first such formalisation
that we are aware of).
The proof that decidable Kuratowski finite sets form a \(\Pi\)-pretopos is
additional to that.
\subsection{Countability Predicates}
After the finite predicates, we will briefly look at the infinite countable
types, and classify them in a parallel way to the finite predicates
(Section~\ref{infinite}).
\todo{Expand}
\subsection{Search}
All of our work is formalised in Cubical Agda
\cite{vezzosiCubicalAgdaDependently2019}: as a result, the constructive
interpretation of each proof is actually a program which can be run on a
computer.
In finiteness in particular, these programs are particularly useful for
exhaustive search.
\todo{Expand}
\section{Notation and Background}
\subsection{Notation}
We work in Cubical Type Theory \cite{cohenCubicalTypeTheory2016}.
For the various type formers we use the following notation:
\begin{description}
  \item[Type] We use \(\mathbf{Type}\) to denote the universe of (small) types.
    ``Type families'' are functions into \(\mathbf{Type}\).
  \item[0 , 1 , 2] We call the \(\mathbf{0}\), \(\mathbf{1}\), and
    \(\mathbf{2}\) types \(\bot\), \(\top\), and \(\mathbf{Bool}\) respectively.
    The single inhabitant of \(\top\) is tt, and the two inhabitants of
    \(\mathbf{Bool}\) are false and true.
    The ``negation'' of a type, written \(\neg A\), means \(A \rightarrow
    \bot\).
  \item[Dependent Sum and Product] We use \(\Sigma\) and \(\Pi\) for the
    dependent sum and product, respectively.
    The two projections from \(\Sigma\) are called fst and snd.
    In the non-dependent case, \(\Sigma\) can be written as \(\times\), and
    \(\Pi\) as \(\rightarrow\).
  \item[Disjoint Union] We define disjoint union as an inductive type.
    \begin{equation}
      \begin{alignedat}{3}
        A \uplus B \coloneqq & \;
        \text{inl} &: A \rightarrow A \uplus B ; \\
        | & \;  \text{inr} &: B \rightarrow A \uplus B ;
      \end{alignedat}
    \end{equation}
    It is also expressible with only \(\Sigma\):
    \(A \uplus B \simeq \Sigma(x : \mathbf{Bool}) , \text{if } x \text{ then
    } A \text{ else } B \).
  \item[Equalities, equivalences, and paths] We use the symbol \(\coloneqq\)
    for definitions.
    \(\simeq\) will be used for equivalences, and \(\equiv\) for equalities.
    Of course, we know that \((A \simeq B) \simeq (A \equiv B)\) by univalence,
    so the distinction isn't terribly important in usage: we will only use one
    or the other as a suggestion of how we constructed it or how it is to be
    used.
  \item[Lambdas] \todo{Figure out how to describe lambdas}
\end{description}
\subsection{Cubical Type Theory}
Cubical Type Theory \cite{cohenCubicalTypeTheory2016} is a constructive type
theory with an implementation in Cubical Agda
\cite{vezzosiCubicalAgdaDependently2019}.
It allows us to do much of the same theory as in HoTT, but crucially the
univalence ``axiom'' is a \emph{theorem}, giving it computational content.
\begin{definition}[Path Types] \label{path-types}
  The equality type (which we denote with \(\equiv\)) in CuTT is the type of
  Paths\footnotemark.
  The internal structure of paths is largely irrelevant to us here, as we will
  generally treat \(\equiv\) as a black-box equivalence relation with
  substitution and congruence.
\end{definition}
